---
output: html_document
---

## Grammar of Graphics {#dv-gg}

The grammar of graphics is a system of rules that describes how data and
graphical aesthetics (e.g. color, size, shape, etc) are combined to form
graphics and plots. First popularized in the book [The Grammar of
Graphics](https://www.google.com/books/edition/_/ZiwLCAAAQBAJ?hl=en) by Leland
Wilkinson and co-authors in 1999, this grammar is a major contribution to the
structural theory of statistical graphics. In 2005, Hadley Wickam wrote an
implementation of the grammar of graphics in R called
[ggplot2](https://ggplot2.tidyverse.org/) (gg stands for grammar of graphics).

Under the grammar of graphics, every plot is the combination of three types of
information: **data**, **geometry**, and **aesthetics**. Data is the data we
wish to plot. Geometry is the type of geometry we wish to use to depict the data
(e.g. circles, squares, lines, etc). Aesthetics connect the data to the geometry
and defines how the data controls the way the selected geometry looks.

A simple example will help to explain. Consider the following made up sample
metadata tibble for a study of subjects who died with Alzheimer's Disease (AD)
and neuropathologically normal controls:

``` {r include=FALSE}
source("0600_data_viz.R")
```

``` {r}
ad_metadata
```

For context, [tau protein](https://en.wikipedia.org/wiki/Tau_protein) and
[amyloid beta peptides from the amyloid precursor
protein](https://en.wikipedia.org/wiki/Amyloid_beta)
aggregate into neurofibrillary tangles and A-beta plaques, respectively, the
brains of people with AD. Generally, the amount of both of these pathologies is
associated with more severe disease. [Braak
stage](https://en.wikipedia.org/wiki/Braak_staging) is a neuropathological
assessment of the amount of pathology in a brain that is associated with the
severity of disease, where 0 indicates absence of pathology and 6 with
widespread involvement in multiple brain regions. Aggregation of tau is also a
[consequence of normal aging](https://pubmed.ncbi.nlm.nih.gov/8866683/), so must
accompany neurological symptoms such as dementia to indicate an AD diagnosis
post mortem. Note we have control samples as well as AD.

![Tauopathy: tau protein accumulates in the cell bodies of affected neurons -
Wikipedia](https://upload.wikimedia.org/wikipedia/commons/f/f7/Tauopathy_in_Alzheimer%27s_disease.jpg)

The histology measures `tau`, `abeta`, `iba1`, and `gfap` have been quantified
using digital microscopy, where brain sections are stained with
immunohistochemistry to identify the location and degree of pathology; the
measures in the table are the number of pixels of a 400 x 400 pixel image of a
piece of brain tissue that fluoresce when stained with the corresponding
antibody. Tau and A-beta antibodies are specialized to the types of aggregated
proteins mentioned above and provide a quantification of the level of overall
AD pathology. Ionized calcium binding adaptor molecule 1 (IBA1) is a marker of
activated [microglia](https://en.wikipedia.org/wiki/Microglia), the resident
macrophages of the brain, which is an
indication of neuroinflammation. Glial fibrillary acidic protein (GFAP) is a
marker for activated [astrocytes](https://en.wikipedia.org/wiki/Astrocyte),
specialized cells that derive from the neuron lineage, are critical for
maintaining the [blood brain
barrier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4292164/), and are also
involved in the neuroinflammatory response.

Let's say we wished to visualize the relationship between age at death and the
amount of tau pathology. A scatter plot where each marker is a subject with $x$
and $y$ position corresponding to `age_at_death` and `tau` respectively. The
following R code creates such a plot with ggplot2:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x = age_at_death, y=tau)) +
  geom_point()
```

All ggplot2 plots begin with the `ggplot()` function call, which is passed a
tibble with the **data** to be plotted. We then define  the **aesthetics** are
defined by mapping the x coordinate to the `age_at_death` column and the y
coordinate to the `tau` column with `aes(x = age_at_death, y=tau)`. Finally, the
**geometry** as 'point' with `geom_point()`, meaning marks will be made at pairs
of x,y coordinates. The plot shows what we expect given our knowledge of the
relationship between age and amount of tau; the two look to be positively
correlated.

However, we are not capturing the whole story: we know that there are both AD
and Control subjects in this dataset. How does `condition` relate to this
relationship we see? We can layer on an additional aesthetic of color to add
this information to the plot:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x = age_at_death, y=tau, color=condition)) +
  geom_point()
```

This looks a little clearer, showing that Control subjects generally have both
an earlier age at death and a lower amount of tau pathology. This might be a
problem, however, since if the age distributions of AD and Control groups are
different that might pose a problem with
[confounding](https://en.wikipedia.org/wiki/Confounding). We should investigate
this.

Instead of plotting age at death and tau against each other, we will examine the
distributions of each of these variables for AD and Control samples separately.
We will use the [violin
geometry](https://ggplot2.tidyverse.org/reference/geom_violin.html) with
`geom_violin()` to look at the distributions of `age_at_death`:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x=condition, y=age_at_death)) +
  geom_violin()
```

We can see immediately that there are big differences between the age
distributions of the two groups. This is not ideal, but perhaps we can adjust
for these effects in downstream analyses. We'd like to look at the tau
distributions as well, but it would be nice to have these two plots side by side
in the same plot. To do that, we will use another library called
[patchwork](https://patchwork.data-imaginist.com/), which allows independent
ggplot2 plots to be arranged together with a simple expressive syntax:

``` {r}
library(patchwork)
age_boxplot <- ggplot(data=ad_metadata, mapping = aes(x=condition, y=age_at_death)) +
  geom_boxplot()
tau_boxplot <- ggplot(data=ad_metadata, mapping=aes(x=condition, y=tau)) +
  geom_boxplot()

age_boxplot | tau_boxplot # this puts the plots side by side
```

This confirms our suspicion, and also reveals a serious problem with our
samples: we have strong confounding of tau and age at death between AD and
Control samples. This means that if we look for differences between AD and
Control, we won't know if the difference is due to the amount of tau pathology
or due to age of the subjects. With this sample set, we simply cannot
confidently answer that question. Just a few simple plots alerted us to this
problem; hopefully more expensive datasets have not already been generated for
these samples, so that hopefully different subjects are available that could
avoid this confounding.

This has been a biological data analysis oriented tutorial on plotting meant to
illustrate the principles of the grammar of graphics. Namely, every plot has
**data**, **geometry**, and **aesthetics** that can be independently controlled
to produce many types of plots. Many of these plots have names, like scatter
plots and boxplots, but as you compose different types of geometries and
aesthetics together you may find yourself generating plots that aren't so easily
named.

The next sections of this chapter are a kind of "cook book" of different kinds
plots you can generate with data of different shapes. It is not intended to be
comprehensive, but a helpful guide when you are trying to decide how to
visualize your own datasets.

::: {.box .note}
If you want to go directly to the comprehensive documentation of the many types
of ggplot2 plots, peruse the [R Graph Gallery](https://www.r-graph-gallery.com/)
site.
:::

::: {.box .readmore}
* [ggplot2 - Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html), by Hadley Wickam
* [R for Data Science - Data Vizualization](https://r4ds.had.co.nz/data-visualisation.html)
* [R Graph Gallery](https://www.r-graph-gallery.com/)
* [patchwork package](https://patchwork.data-imaginist.com/)
:::

### `ggplot` mechanics

`ggplot` has two key concepts that give it great flexibility: **layers** and
**scales*.

Every plot has one or more layers that contain a type of geometry that
represents a data encoding. In general, each layer will only have one geometry
type, e.g. points or lines, but the geometry might be complex, e.g. density
plots. The layers added to a plot form a stack, where the layers added first are
beneath those added later. The geometry in each layer may draw from the same
data, or each may have its own. Each layer may also share the aesthetic mapping
from the `ggplot()` call, or may have its own. This is why both the `ggplot()`
function and each individual `geom_X()`  function can accept data and aesthetic
mappings. The package comes with a large number of geometries described in its
[reference documentation](https://ggplot2.tidyverse.org/reference/index.html#geoms).

The geometry in each layer maps the data values to visual properties using
scales. A scale may map a data range to a pixel range, or to a color on a color
gradient, or one of a set of discrete colors or shapes. `ggplot` provides
reasonable default scales for each geometry type. You can override these
defaults by using the [`scale_X`
functions](https://ggplot2.tidyverse.org/reference/index.html#scales).

The [ggplot2 book](https://ggplot2-book.org/) is an excellent resource for all
things ggplot2.