---
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
```

## Plotting One Dimension

The simplest plots involve plotting a single vector of numbers, or several such
vectors (e.g. for different samples). Each value in the vector typically
corresponds to a category or fixed value, for example the `tau` column from the
example above has pairs of *(ID, tau value)*. The order of these numbers can be
changed, but the vector remains one dimensional or 1-D.

### Bar chart {#dv-bar}

Bar charts map length (i.e. the height or width of a box) to the scalar value of
a number. The difference in visual length can help the viewer notice consistent
patterns in groups of bars, depending on how they are arranged:

``` {r}
ggplot(ad_metadata, mapping = aes(x=ID,y=tau)) +
  geom_bar(stat="identity")
```

Note the `stat="identity"` argument is required because by default `geom_bar`
counts the number of values for each value of x, which in our case is only ever
one. This plot is not particularly helpful, so let's change the fill color of
the bars based on condition:

``` {r}
ggplot(ad_metadata, mapping = aes(x=ID,y=tau,fill=condition)) +
  geom_bar(stat="identity")
```

Slightly better, but maybe we can see even more clearly if we sort our tibble
by tau first. Sorting elements in these 1-D charts is somewhat complicated, and
is explained in the [Reordering 1-D Data Elements] section below.

Bar charts can also plot negative numbers. In the following example, we center
the tau measurements by subtracting the mean from each value before plotting:

``` {r}
mutate(ad_metadata, tau_centered=(tau - mean(tau))) %>%
  ggplot(mapping = aes(x=ID, y=tau_centered, fill=condition)) +
  geom_bar(stat="identity")
```

### Lollipop plots

Similar to [bar charts](#dv-bar), so-called "lollipop plots" replace the bar
with a line segment and a circle. The length of the line segment is proportional
to the magnitude of the number, and the point marks the length of the segment as
a height on the y or length on the x axis, depending on orientation.

``` {r}
ggplot(ad_metadata) +
  geom_point(mapping=aes(x=ID, y=tau)) +
  geom_segment(mapping=aes(x=ID, xend=ID, y=0, yend=tau))
```

Note that `aes()` mappings can be made on the `ggplot()` object or on each
individual geometry function call, to specify different mappings based on
geometry.

### Stacked Area charts

Stacked area charts can visualize multiple 1D data that share a common
categorical axis. The charts consist of one line per variable with vertices that
correspond to x and y values similar to a [bar](#bar-plots) or
[lollipop](#lollipop-plots) plots. Each variable is plotted using the previous
one as a baseline, so that the height of the data points for each category is
proportional to their sum. The space between the lines for each variable and the
previous one are filled with a color. The following plot visualizes the amount
of marker stain for each of the four genes for each individaul:

```{r}
pivot_longer(
  ad_metadata,
  c(tau,abeta,iba1,gfap),
  names_to='Marker',
  values_to='Intensity'
  ) %>%
  ggplot(aes(x=ID,y=Intensity,group=Marker,fill=Marker)) +
    geom_area()
```

We notice that subject A4 has the highest overall level of marker intensity,
followed by A1, A7, etc. The control samples overall have less intensity across
all markers. Certain samples, A2 and C5, have little to no `abeta` aggregation,
and C6 has little to no `tau`.

Stacked area plots require three pieces of data:

* **x** - a numeric or categorical axis for vertical alignment
* **y** - a numeric axis to draw vertical proportions
* **group** - a categorical variable that indicates which (x,y) pairs correspond
to the same line

In the example above, we needed to pivot our tibble so that the different
markers and their values were placed into columns `Marker` and `Intensity`,
respectively. Data for stacked bar charts will usually need to be in this 'long'
format, as described in [Rearranging Data].

Sometimes it is more helpful to view the relative proportion of values in each
category rather than the actual values. The result is called a *proportional*
stacked area plots. While not a distinct plot type, we can create one by
preprocessing our data by dividing each value by the column sum:


```{r}
pivot_longer(
  ad_metadata,
  c(tau,abeta,iba1,gfap),
  names_to='Marker',
  values_to='Intensity'
  ) %>%
  group_by(ID) %>% # we want to divide each subjects intensity values by the sum of all four markers
  mutate(
    `Relative Intensity`=Intensity/sum(Intensity)
  ) %>%
  ungroup() %>% # ungroup restores the tibble to its original number of rows after the transformation
  ggplot(aes(x=ID,y=`Relative Intensity`,group=Marker,fill=Marker)) +
    geom_area()
```
Now the values for each subject have been normalized to each sum to 1. In this
way, we might note that the relative proportion of `abeta` seems to be greater
in AD samples than Controls, but that may not be true of `tau`. These
observations may inspire us to ask these questions more rigorously than we have
done so far by inspection.