---
output: html_document
---

## Visualizing Distributions

The distribution is one of the most important properties of a set of numbers.
A distribution describes the general "shape" of the numbers, i.e. what is the
relative frequency of the values, or ranges of values, within the data.
Understanding the distribution of a data set is critical when choosing methods
to apply, since many methods are only appropriate when data is distributed in
certain ways, e.g. [linear
regression](https://en.wikipedia.org/wiki/Linear_regression) assumes the
response variable is [normally
distributed](https://en.wikipedia.org/wiki/Normal_distribution), otherwise the
result of the model cannot be interpreted properly. Often, we don't know how our
data are distributed when we obtain it and so we must examine the distribution
empirically. The visualizations in this section are all used for the purpose of
depicting the distribution of a set of numbers.

### Histogram

The most common way to plot the distribution of a 1-D set of data is the
[histogram](https://en.wikipedia.org/wiki/Histogram). The histogram divides up
the range of a dataset from minimum to maximum into bins usually of the same
width and tabulates the number of values that fall within each bin. Below is a
histogram of our `age_at_death` measurement for all samples:

``` {r}
ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death))
```

Note that the histogram does not look very complete, because there are only 20
values in our data. We can mitigate this somewhat by increasing the number of
bins the data range is divided into:

``` {r}
ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death),bins=10)
```

This is a little bit better, but there are still some bins (76-79, 84-87) that
have no values. Compare this to the following synthetic dataset of 1000
normally distributed values:

``` {r}
tibble(
  x=rnorm(1000)
) %>%
  ggplot() +
  geom_histogram(aes(x=x))
```

For distributions with a small number of samples, histograms might not be the
best visualization. We will continue with synthetic normally distributed dataset
for the remaining examples.

ggplot allows you to easily plot multiple distributions on the same plot:

``` {r}
tibble(
  x=c(rnorm(1000),rnorm(1000,mean=4)),
  type=c(rep('A',1000),rep('B',1000))
) %>%
  ggplot(aes(x=x,fill=type)) +
  geom_histogram(bins=30, alpha=0.6, position="identity")
```

The `alpha=0.6, position="identity"` arguments makes the bars partially
transparent so you can see the overlap more clearly.

::: {.box .readmore}
* [ggplot2 geom_histogram reference](https://ggplot2.tidyverse.org/reference/geom_histogram.html?q=histogram)
* [R Graph Gallery - histograms](https://www.r-graph-gallery.com/histogram.html)
:::

### Density

Another way to describe a distribution is with a [density
plot](https://www.r-graph-gallery.com/density-plot.html). Instead of binning the
values into intervals and drawing bars with height proportional to the number of
values in each bin, a density plot draws a smoothly interpolated line that
approximates the distribution instead. A key difference between a histogram and
a density plot is the density plot is always normalized so the integral under
the curve is approximately 1, whereas a histogram may be either counts or, if
the counts in each bin are divided by the total number of data points, a
proportion.

Compare the histogram and density plots
of the  `age_at_death` variable from our example tibble:

``` {r}
library(patchwork)
hist_g <- ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death),bins=30)
density_g <- ggplot(ad_metadata) +
  geom_density(mapping=aes(x=age_at_death),fill="#c9a13daa")

hist_g | density_g
```

Notice the overall shape of the two distributions is similar, with the highest
values in both around age 77. The density plot is a smoother representation of
a histogram, but its accuracy is still highly sensitive to the number of
measurements used to construct it. Compare the histogram and density plots of
two sets of 1000 normally distributed samples with different means:

``` {r}
library(patchwork)
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,mean=4)),
  type=c(rep('A',1000),rep('B',1000))
)
hist_g <- ggplot(normal_samples) +
  geom_histogram(
    mapping=aes(x=x,fill=type),
    alpha=0.6,
    position="identity",
    bins=30
)
density_g <- ggplot(normal_samples) +
  geom_density(
    mapping=aes(x=x,fill=type),
    alpha=0.6,
    position="identity"
  )

hist_g | density_g
```

Again the two types of plots depict similar distributions, although they are
different enough to possibly suggest different interpretations. In general,
density plots might be preferable over histograms if the data are noisy or
sparse in that they produce cleaner plots, but potentially at the expense of
accuracy when the number of samples is low.

::: {.box .readmore}
* [ggplot2 geom_density reference](https://ggplot2.tidyverse.org/reference/geom_density.html?q=density)
* [R Graph Gallery - density plots](https://www.r-graph-gallery.com/density-plot.html)
:::

### Boxplot

Box plots, or box and whisker plots, are extremely common when used to describe
distributions. Below is a boxplot of age at death divided by condition:

``` {r}
ggplot(ad_metadata) +
  geom_boxplot(mapping=aes(x=condition,y=age_at_death))
```

Boxplots are drawn assuming the data are
[unimodal](https://en.wikipedia.org/wiki/Unimodality) (i.e. shaped like a hill,
possibly slanted to one side or the other), where the extents of the box
represent the 1st and 3rd quartile of the data, the central line is the median,
the whiskers are drawn as 1.5 times the value outside the 1st and 3rd quartiles.
Sometimes individual values more extreme than the whiskers are drawn
individually to identify them as outliers.

![Boxplot anatomy. IQR stands for "inner quartile range", the distance between
the 1st and 3rd quartile - Wikipedia](https://upload.wikimedia.org/wikipedia/commons/1/1a/Boxplot_vs_PDF.svg)

However, boxplots have some significant shortcomings. Primarily, the rectangle
of the inner quartile range does not describe the actual distribution of the
samples within it. Although the median can give a sense of skewness, if the
data are not unimodal this may be misleading. Consider the following
distributions plotted as boxplots or as violin plots (described in the next
section):

``` {r}
library(patchwork)
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,4),rnorm(1000,2,3)),
  type=c(rep('A',2000),rep('B',1000))
)
g <- ggplot(normal_samples, aes(x=type,y=x,fill=type))
boxplot_g <- g +  geom_boxplot()
violin_g <- g + geom_violin()

boxplot_g | violin_g
```

The two distributions look almost identical in the boxplot figure; however they
are dramatically different when visualized using a method like a violin plot
where the contours of the entire distribution are depicted. Unless you are
certain that your data are unimodal, one of the other distribution visualization
methods in this section will likely more accurately depict your data than a
boxplot.

::: {.box .readmore}
* [ggplot2 geom_boxplot reference](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)
* [R Graph Gallery - boxplots](https://www.r-graph-gallery.com/boxplot.html)
:::

### Violin plot

As seen in the last section, a [violin
plot](https://ggplot2.tidyverse.org/reference/geom_violin.html) is another way
to depict a distribution by producing a shape where the width is proportional to
the value along the x or y axis, depending on orientation. The "violin" shape is
similar in principle to a histogram or a density plot, in that it describes the
contour of all the data in the distribution, not just the quantiles and extents,
as in a box plot. Below is a violin plot of the `tau` measures from our example
tibble:

``` {r}
ggplot(ad_metadata) +
  geom_violin(aes(x=condition,y=tau,fill=condition))
```

The violin plot is both more and less descriptive than a boxplot; it does depict
the entire distribution of the data, but also doesn't include features like
median by default.

::: {.box .readmore}
* [ggplot2 geom_violin reference](https://ggplot2.tidyverse.org/reference/geom_violin.html)
* [R Graph Gallery - violin plots](https://www.r-graph-gallery.com/violin.html)
:::

### Beeswarm plot

The [beeswarm plot](https://r-charts.com/distribution/ggbeeswarm/) is similar to
a violin plot, but instead of plotting the contours of the data, it plots the
data itself as points like in a scatter plot. The individual values of the
distribution are organized vertically and spaced such that the points don't
overlap. In this plot, the distribution of age at death is plotted for each kind
of sample and the markers are colored by the amount of tau:

``` {r}
library(ggbeeswarm)
ggplot(ad_metadata) +
  geom_beeswarm(aes(x=condition,y=age_at_death,color=condition),cex=2,size=2)
```

We may not have noticed before that our AD samples have a big gap in ages
between 74 and 81; since the beeswarm plot displays all the data, we can see it
easily here.

Beeswarm plots are typically only useful when the number of values is within a
range; not too many and not too few. The example above is close to having too
few values per group for this plot to be useful, but consider the following with
too many samples:

``` {r}
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,4),rnorm(1000,2,3)),
  type=c(rep('A',2000),rep('B',1000))
)
ggplot(normal_samples, aes(x=type,y=x,color=type)) +
  geom_beeswarm()
```

This plot likely has too many samples to be the right choice (it's also ugly),
but it does give an idea of the distribution of the data.

In the previous examples the markers for each group also determined the color of
the group. This makes the chart a bit easier to read and more pleasing to the
eye, but is technically redundant. You can use however profitably however to
color markers by some other value that might be of interest. Consider this final
example where markers are colored by another randomly generated variable:

``` {r}
normal_samples <- tibble(
  x=c(rnorm(100),rnorm(100,4),rnorm(100,2,3)),
  type=c(rep('A',200),rep('B',100)),
  category=sample(c('healthy','disease'),300,replace=TRUE)
)
ggplot(normal_samples, aes(x=type,y=x,color=category)) +
  geom_beeswarm()
```

We are now effectively visualizing three dimensions which may provide insight
into the data.

::: {.box .readmore}
* [beeswarm package reference](https://r-charts.com/distribution/ggbeeswarm/)
:::

### Ridgeline

If you have many non-trivial distributions that you would like the user to
compare, a good option is a [ridgeline
chart](https://www.r-graph-gallery.com/294-basic-ridgeline-plot.html). The
ridgeline plot is simply multiple density plots drawn for different variables
within the same plot. Like the beeswarm plot, ridgeline plots are provided by
another package outside ggplot2.

``` {r}
library(ggridges)

tibble(
  x=c(rnorm(100),rnorm(100,4),rnorm(100,2,3)),
  type=c(rep('A',200),rep('B',100)),
) %>%
  ggplot(aes(y=type,x=x,fill=type)) +
  geom_density_ridges()
```

Many distributions may be plotted:

``` {r}
tibble(
  x=rnorm(10000,mean=runif(10,1,10),sd=runif(2,1,4)),
  type=rep(c("A","B","C","D","E","F","G","H","I","J"),1000)
) %>%
  ggplot(aes(y=type,x=x,fill=type)) +
  geom_density_ridges(alpha=0.6,position="identity")
```

::: {.box .readmore}
* [R Graph Gallery - ridgeline plots](https://www.r-graph-gallery.com/294-basic-ridgeline-plot.html)
* [ggridges package vignette](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html)
* [ggridges package index on CRAN](https://cran.r-project.org/web/packages/ggridges/index.html)
:::
