---
output: html_document
---


## Multiple Plots

Sometimes it is convenient to plot multiple sets of data at once. This can be
accomplished in two ways: facet wrapping to separate out different subsets of a
dataset into plots with shared axes, and adding multiple plots to the same
figure.

### Facet wrapping

Facet wrapping separates subsets of a dataset into plots with identical axes.
This can be useful if different groups of data points overlap and occlude one
another. Consider the following dataset with three groups of normally
distributed random samples:

```{r viz facet oneplot}
library(mvtnorm) # package implementing multivariate normal distributions
nsamp <- 100
data <-  rbind(
    rmvnorm(nsamp,c(1,1),sigma=matrix(c(1,0.8,0.8,1),nrow=2)),
    rmvnorm(nsamp,c(1,1),sigma=matrix(c(1,-0.8,-0.8,1),nrow=2)),
    rmvnorm(nsamp,c(1,1),sigma=matrix(c(1,0,0,1),nrow=2))
)
colnames(data) <- c('x','y')
g_oneplot <- as_tibble(data) %>%
  mutate(
    sample_name=c(rep('A',nsamp),rep('B',nsamp),rep('C',nsamp))
  ) %>%
  ggplot(aes(x=x,y=y,color=sample_name)) +
  geom_point()
g_oneplot
```

The three sample data points are plotted on top of each other, and although they
are colored differently, it is difficult to discern the patterns of each group
clearly. This same plot may be split into three, one for each sample group,
using the [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html)
function:

```{r viz facet wrap}
g_oneplot + facet_wrap(vars(sample_name))
```

Now the trends of each group are clearly visible. The `facet_wrap()` function
accepts one required argument, which is the variable name in the data that
should be used to separate the groups decorated by the `vars()` function. Any
geometry may be used with a facet wrap. For example, we could add a
`geom_smooth()` to the plot to estimate a smoothing function within each facet:

```{r viz facet wrap plus}
g_oneplot + facet_wrap(vars(sample_name)) +
  geom_smooth(method="loess", formula=y ~ x)
```

By default, facet wrapping fixes all the axes of each plot to be identical,
which enables the most accurate estimates of comparisons between plots according
to the [Elementary Perceptual Tasks] hierarchy. This and many other aspects of
the faceting may be adjusted using arguments as described in the
[documentation](https://ggplot2.tidyverse.org/reference/facet_wrap.html).

### Multipanel Figures

Facet wrapping is useful when all of the data for the separate plots are on the
same scale, and comparisons of the facet groups is meaningful. However, faceting
is less convenient when we desire to place plots with unrelated data in the same
figure. The [`patchwork`](https://patchwork.data-imaginist.com/) package allows
us to compose multiple plots together using an intuitive set of operators. For
example, we may put two plots next to each other by separating them with the `|`
operator:

```{r viz patch simple}
data <- tibble(
  a=rnorm(100,0,1),
  b=rnorm(100,3,2)
)
g_scatter <- ggplot(data, aes(x=a, y=b)) +
  geom_point()
g_violin <- pivot_longer(data, c(a,b)) %>% ggplot(aes(x=name,y=value,fill=name)) +
  geom_violin()

g_scatter | g_violin
```

Note that each plot was saved to its own variable and then composed with the
patchwork operator `|`, which puts the plots side-by-side. We could also place
one plot on top of the other with the `/` operator:

```{r viz patchwork vert}
g_scatter / g_violin
```

Plots may be composed into arbitrarily complex multipanel figures in this way
using `()` to group plots together:

```{r patch complex}
g_scatter / ( g_scatter | g_violin)
```

Another possible composition:

```{r viz complex 2}
(g_scatter / g_scatter ) | g_violin
```

The patchwork library is very flexible, and allows the designer to compose plots
in many different ways. These multipanel plots make excellent starting points
for producing [publication-quality figures](publication-ready-plots), as
described in the next section.