---
output: html_document
---

## Network visualization 

The most commonly used package to visualize network is `igraph`.  
```{r message=FALSE, warning=FALSE}
library(igraph)
```

First, let's create a graph with 15 nodes.  

```{r}
names <- c("Opal", "Tourmaline", "Emerald", "Citrine", "Jasper", "Moonstone", "Agate", "Chalcedony", "Crystal", "Obsidian", "Quartz", "Topaz", "Beryl", "Sugilite", "Sapphire")
```

Use `sample_gnm()` to generate a random graph.  
```{r}
set.seed(10)
samp_g <- sample_gnm(
  n = 15, # number of vertices
  m = 20, # total number of edges
  directed = FALSE # if it's a directed network
)
```

To assign color to vertices, use `vertex_attr()`. `gorder()` returns the number of vertices.  
```{r}
vertex_attr(samp_g) <- list(name = names, color = rep("slategray1", gorder(samp_g)))
```

To assign color to edges, use `vertex_attr()`. `gsize()` returns the total number of edges. Here I am assigning random colors. 
```{r}
edge_attr(samp_g) <- list(color = sample(
  x = c("slateblue1", "lightpink1", "steelblue1", "palegreen3"),
  size = gsize(samp_g), replace = TRUE
))
```

In real case, when you want to modify the color of a specific edge, you need to first find which edge you want to modify...
```{r}
edges(samp_g)
```

Then modify it. For example, modify the 1st edge, which is "Emerald--Citrine".  
```{r}
edge_attr(samp_g, name = "color")[1] <- "red"
```

To visualize the network, just use `plot()`!  
```{r}
plot(samp_g)
```

In real-world, we don't want a random network. Now let's create our network from scratch.  

There are many different ways to create a network. One of the most basic (and easy) ways is using `graph_from_data_frame()` function in `igraph`.  
1. Create 2 columns in the data frame: `from` and `to`. Each row represents an edge, and the entries are vertex names. For instance, as indicated in row 1 below, there is an edge "from Crystal to Moonstone". In an undirected graph, you can put the two vertices in either way.  
2. (optional) Add another column `color` to represent the color to plot the edge.  
3. (optional) You can also add additional information for each edge. For example, add a column `weight` to specify the weight of this edge. The `weight` will become useful when you do network analysis later, for example, find the shortest path between two vertices. You can also add other attributes you want, like the "connection_type" here. 
```{r}
set.seed(4)
my_gem_df <- data.frame(
  "from" = c("Crystal", "Crystal", "Crystal", "Crystal", "Topaz", "Obsidian", "Agate", "Topaz"),
  "to" = c("Moonstone", "Citrine", "Obsidian", "Agate", "Citrine", "Topaz", "Moonstone", "Moonstone"),
  "color" = c("slateblue1", "slateblue1", "slateblue1", "steelblue1", "palegreen3", "steelblue1", "lightpink1", "palegreen3"),
  "connection_type" = c(rep(1, 6), 2, 2)
)
my_gem_df$weight <- as.numeric(as.factor(my_gem_df$color))
my_gem_df
```

Now create an igraph object from the data frame, and assign vertex color. 
```{r}
g <- graph_from_data_frame(my_gem_df, directed = FALSE)
vertex_attr(g) <- list(
  name = c("Crystal", "Topaz", "Obsidian", "Agate", "Moonstone", "Citrine"),
  color = c("slategray1", "slategray1", "slategray1", "lightpink", "slategray1", "lightpink")
)
```


Again, simply use `plot(g)` to plot the network. We can also do some fancy customization to play around with the labels! *(Remember the `connection_type` I specified above in the data frame? In `edge.lty`, 1 is a solid line, and 2 is a dash line)* 
```{r}
set.seed(4)
plot(g,
  edge.width = 1.5,
  edge.lty = edge_attr(g, "connection_type"),
  vertex.label.family = "Trebuchet MS",
  vertex.label.font = 3,
  vertex.label.cex = 1,
  vertex.label.dist = 2,
  vertex.label.degree = pi / 2
)
```

To see the whole list of available arguments available when plotting your network, visit [igraph plot manual](https://igraph.org/r/doc/plot.common.html).  

In the case of an directed graph, simply put `directed = TRUE` when creating the igraph object, and you will see the cute arrows.  
```{r}
g <- graph_from_data_frame(my_gem_df, directed = TRUE)
vertex_attr(g) <- list(
  name = c("Crystal", "Topaz", "Obsidian", "Agate", "Moonstone", "Citrine"),
  color = c("slategray1", "slategray1", "slategray1", "lightpink", "slategray1", "lightpink")
)
set.seed(4)
plot(g,
  edge.width = 1.5,
  edge.lty = edge_attr(g, "connection_type"),
  edge.arrow.size = 0.8,
  vertex.label.family = "Trebuchet MS",
  vertex.label.font = 3,
  vertex.label.cex = 1,
  vertex.label.dist = 2,
  vertex.label.degree = pi / 2
)
```

### Interactive network plots 
Besides `igraph`, there are many other packages to create fancier networks. One example is `visNetwork`. It can create an interactive network, and it's easy to use! Check [example here](https://cran.r-project.org/web/packages/visNetwork/vignettes/Introduction-to-visNetwork.html). It can be implemented in your HTML report knitted from R markdown and R Shiny app interface!  
One thing to keep in mind: as you can imagine, since it is interactive, `visNetwork` requires much more memory and time compared to a static graph. So, you may need to make a choice when you develop your web interface or R markdown HTML report.  
