---
output: html_document
---

```{r, include=FALSE}
source("0700_bioinfo.R")
```

## Gene Set Enrichment Analysis

With the constant evolution of high-throughput sequencing (HTS) technologies,
the size and dimensionality of data generated has been ever increasing. The
question of interest has shifted from how do we generate the data to how do we
make meaningful biological interpretations on a genome-wide level. The simplest
and most common output of HTS experiments is a list of "interesting" genes. In
the specific case of differential gene expression analysis, it is possible and
quite common to obtain hundreds or even thousands of differentially expressed
genes in a single experiment that may be directly or indirectly related to the
phenotype of interest.

While it can be helpful and fruitful to research these genes individually, this
form of personal inspection is limited by one's domain knowledge and by the size
of the results. On a biological level, it is complicated by the fact that most
processes are often coordinated by the actions of many genes/gene products
working in concert. Cellular signaling pathways, phenotypic differences or
responses to various stimuli are typically associated with changes in the
expression pattern of many genes that share common biological functions or
regulation.

Gene set enrichment analysis is an umbrella term for methods designed to analyze
expression data and capture changes in higher-level biological processes and
pathways by organizing genes into biologically relevant groups or gene sets. We
will discuss the background of two common gene set enrichment analyses
(over-representation analysis and rank-based Gene Set Enrichment Analysis),
their advantages and disadvantages, and walk-through an example of how they can
each be implemented in R. Before this, we will briefly touch upon the definition
of a gene set and describe how to construct or obtain gene sets.

### Gene Sets

Gene sets are curated lists of genes that are grouped together by some logical
association or pre-existing domain knowledge. Their primary use is to facilitate
the biological interpretation of expression data by capturing high-level
interactions between biologically relevant groups of genes. These sets are
highly flexible and may be constructed based on any *a priori* knowledge or
classification. For example, one could define a gene set that includes all genes
that are members of the same biochemical pathway or a gene set that consists of
all genes that are induced upon treatment with a particular pharmacological
agent. While it is perfectly valid to construct new gene sets, there are many
existing and curated collections of gene sets that are maintained and
contributed to by communities of scientists throughout the world. Below, we will
highlight some of the major collections that are commonly used and cited in
published work:

__[KEGG Pathways](https://www.genome.jp/kegg/pathway.html)__

The Kyoto Encyclopedia of Genes and Genomes (KEGG) is a repository for
biological pathway information that the authors describe as a means to
"computerize functional interpretations as part of the pathway reconstruction
process based on the hierarchically structured knowledge about the genomic,
chemical, and network spaces". The KEGG Pathways database consists of maps
displaying the functional and regulatory relationships of genes within various
metabolic and cell signaling pathways. The KEGG offers a web service that allows
for the extraction of the genes and other information belonging to specific
pathways.

__[GO Annotations](http://geneontology.org/docs/go-annotations/)__

The Gene Ontology provides a network representation of biological systems from
the molecular-level to the organismal level by defining a graph based
representation of connected terms using a controlled vocabulary. At a high
level, GO annotations consist of three broad ontologies termed molecular
function, cellular component, and biological process. The molecular function is
the specific activity a gene product performs on a molecular level (i.e. a
protein kinase would be annotated with *protein kinase activity*). The
biological process defines the higher-level programs and pathways that's
accomplished by the activities of the gene product (i.e. if we take our previous
example of a protein kinase, it's biological process might be assigned as
*signal transduction*). The cellular component refers to the localization of the
gene product within the cell (i.e. the cellular component of a protein kinase
might be *cytosol*). A gene product can be annotated with zero or more terms
from each ontology, and these annotations are based on multiple levels of
evidence from published work. The GO annotations may be accessed or downloaded
directly from their own website, and there exist a number of web services that
use GO annotations in the background ([DAVID](https://david.ncifcrf.gov),
[enrichR](https://maayanlab.cloud/Enrichr/), etc.)

__[Molecular Signatures Database](http://www.gsea-msigdb.org/gsea/msigdb/)__

The Molecular Signatures Database (MSigDB) is a collection of gene sets curated,
maintained and provided by the Broad Institute and UC San Diego. Although
intended and designed for specific use with the Gene Set Enrichment Analysis
Methodology, they are freely available and only require proper attribution for
other uses. The MSigDB consists of 9 major collections of gene sets: H (hallmark
gene sets), C1 (positional gene sets), C2 (curated gene sets), c3 (regulatory
target gene sets), c4 (computational gene sets), c5 (ontology gene sets), c6
(oncogenic signature gene sets), c7 (immunologic signature gene sets), c8 (cell
type signature gene sets). They are available in formats ready for use in the
GSEA methodology and other formats that are easily imported into various
settings for custom use. The gene sets as well as the GSEA methodology are
available directly from their website. There are a number of R-specific packages
that have been developed for working directly with these gene sets such as
[GSEABase](https://bioconductor.org/packages/release/bioc/html/GSEABase.html) or
[fgsea](http://bioconductor.org/packages/release/bioc/html/fgsea.html), which we
will discuss later.

### Working with gene sets in R

We will walk through two quick examples of how to read in an example collection
of gene sets. We will utilize the Hallmarks gene set collection, downloaded
directly from the MSigDB website, which consists of 50 gene sets representing
well-defined biological processes and generated by aggregating together many
pre-existing gene sets. The MSigDB provides these gene set collections in the
GMT format. These files are tab-delimited and each row in the format represents
one gene set with the first column being the name of the gene set, and the
second column a short description. The remaining columns each represent a gene
and unequal lengths of columns per row is allowed. As is, the GMT format is
designed to work specifically with the GSEA methodology developed and provided
by the Broad Institute. However, we will also show two ways to manually parse
these gene collections for exploration and further use in R.

The first way to parse these gene sets would be to use various tidyverse
functions that you should be familiar with already to construct a tibble.
Essentially, we read the file, rename the first two columns for convenience, and
use a combination of `pivot_longer()` and `group_by()` to quickly access genes
by pathway. Below, we have the results of these operations and used them to
display a summary of the number of genes in all the pathways contained with the
hallmark pathways gene collection.

```{r read in GMT file using tidyverse, warning=FALSE, message=FALSE}
read_delim('h.all.v7.5.1.symbols.gmt', delim='\t', col_names=FALSE) %>%
  dplyr::rename(Pathway = X1, Desc=X2) %>%
  dplyr::select(-Desc) %>%
  pivot_longer(!Pathway, values_to='genes', values_drop_na=TRUE, names_to = NULL) %>%
  group_by(Pathway) %>%
  summarise(n=n())
```
We can see that there are 50 total pathways in the hallmarks gene collection
that contain varying numbers of genes. We discarded the description column and
if we were to save the results after using `pivot_longer()` instead of piping
them to `group_by` and `summarise()`, we would have a tibble in the "long"
format with each row representing a single pathway and single gene.

However, there do exist various packages that have been developed to
specifically handle gene set data in R such as the previously mentioned
GSEABase. This is a collection of functions and class-based objects that
facilitate working with gene sets. The foundation of the GSEABase package is the
GeneSet and GeneSetCollection classes which store gene sets and metadata or a
collection of GeneSet objects, respectively. We will use the GSEABase package to
read in the collection of gene sets we previously downloaded.

```{r, include=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GSEABase")
```

```{r GSEABase read and access GMT data, warning=FALSE, message=FALSE}
library('GSEABase')
hallmarks_gmt <- getGmt(con='h.all.v7.5.1.symbols.gmt')
hallmarks_gmt
```
If we simply access the hallmarks_gmt variable, we can see that it is a
GeneSetCollection object containing `r length(names(hallmarks_gmt))` total gene
sets which encompass 4383 unique identifiers or HGNC symbols. Although this
package supports a range of functions, we will focus on the basics. For a more
thorough description of the classes and methods, please read their extended
documentation available
[here](https://bioconductor.org/packages/release/bioc/manuals/GSEABase/man/GSEABase.pdf).

The geneIds method will return a list with each pathway as a named vector of associated gene ids:
```{r GeneSetCollection method geneIds}
head(geneIds(hallmarks_gmt), 2)
```

The names method will return all of the gene set names contained within a specific collection:
```{r GeneSetCollection method names}
head(names(hallmarks_gmt),2)

```

We can access a specific gene set contained within this collection by referring
to its name and using the following notation:

```{r access the Adipogenesis gene set}
hallmarks_gmt[['HALLMARK_ANGIOGENESIS']]
```
Simply accessing the object will provide a high-level summary of the information
contained within. To access a specific value of this GeneSet object, we would
call one of the slots (a core concept in object-oriented programming). In our
particular case, we could extract the gene names contained assigned to this
GeneSet by calling the `geneIds` slot as shown below to return a vector of the
gene names. We can see the first five below and also the length of the returned
vector by using the base R `length()` function:

```{r sample geneIds slot}
head(hallmarks_gmt[['HALLMARK_ANGIOGENESIS']]@geneIds)
```

```{r sample geneIds length}
length(hallmarks_gmt[['HALLMARK_ANGIOGENESIS']]@geneIds)
```

GSEABase includes a number of built-in functions for reading gene sets in from
various sources and performing common operations such as set intersections, set
differences, and ID conversions. We will demonstrate the usage of some of these
in the next section covering Over-representation analysis.

### Over-representation Analysis

One of the most common ways to utilize gene sets to evaluate gene expression
results is to perform Over-representation Analysis (ORA). Let us assume that we
have obtained a list of differentially expressed genes from an experiment. We
are curious to know if within this list of differentially expressed genes, do we
see an *over-representation* or *enrichment* of genes belonging to gene sets of
interest? In more general terms, the goal of ORA is to determine how likely it
is that there is a non-random association between a gene being differentially
expressed and having membership in a chosen (and ideally relevant) gene set. In
R, we can do a simple ORA by utilizing a Fisher's exact test and a contingency
table.

For a purely hypothetical example, let us assume that we have performed
differential gene expression analysis between two different cell lines. We
obtain a list of 10,000 total genes (our background) discovered in the
experiment and find that at our chosen statistical threshold, 1,000 of these are
differentially expressed. To keep things simple, we will perform a single ORA
test against the Hallmarks Angiogenesis gene set using a sample list of 1000
differentially expressed genes selected from data generated by [Marisa et al.
2013](https://pubmed.ncbi.nlm.nih.gov/23700391/). The Hallmarks Angiogenesis
gene set consists of 36 genes and we find that 13 of these are also present in
our list of differentially expressed genes.

:::{.box .warning}
Please note that we are removing this list of genes from its original meaning
and context found in the publication and simply using it to demonstrate the
basic steps occurring during ORA. All of these numbers and lists were
arbitrarily chosen and this experimental setup is purely hypothetical.

Another issue to note is that typically the "background" list should represent
the entire pool of genes from which any differentially expressed genes could
have been selected. For expression experiments, it is typical to choose all of
the detected genes (regardless of significance) as the background. The number of
genes in the organism's genome could potentially also be an appropriate
"background"   

Also, it is important to keep in mind that in reality, ORA is nearly always
performed on a larger scale against a variety of different gene sets. This
allows for the unbiased discovery of potentially novel and unexpected enrichment
in other biological areas of interest. It also necessitates the need for
multiple-testing correction, which we have discussed in [multiple hypothesis
testing](#multiple-hypothesis-testing).
:::

To begin, we would want to prepare a contingency table which describes the
various overlaps between our sets of interest. For a 2x2 contingency table,
these four values will be:

1. Genes present in our list of differentially expressed genes and present in our gene set
2. Genes present in our list of differentially expressed genes and not present in our gene set
3. Genes not present in our list of differentially expressed genes and present in our gene set
4. Genes not present in our list of differentially expressed genes and not present in our gene set

To demonstrate what this would look like, we have manually constructed a
contingency table with labels and totals added below. If you look at the margins
of the table and recall the previously given values above, you can reconstruct
the logic used to generate each of the values in all the cells.

```{r example contingency table, echo=FALSE}
Total <- sum
row_names = c('In Gene Set', 'Not in Gene Set')
col_names = c('Differentially Expressed', 'Not Differentially Expressed')

knitr::kable(addmargins(matrix(c(13, 987, 23, 8977), nrow=2, dimnames=list(row_names, col_names)), FUN=Total, quiet=TRUE))
```

For the purposes of this example, we are reading in our differentially expressed
genes from an external file, but this vector could be generated in any number of
ways depending upon where and how your results are stored. Following good coding
practices, we will write a small function that takes this list of DE genes and a
GeneSet object to programmatically generate a contingency table:

:::{.box .note}
There are many ways to construct a contingency table. This is just one way that
was chosen to make calculations of the values contained within the table
transparent and easy to understand.
:::

```{r mock function to make contingency table}
#load and read our list of DE genes contained within a newline delimited txt file
de_genes <- scan('example_de_list.txt', what=character(), sep='\n')

#define a function that takes a list of DE genes, and a specific GeneSet from a GeneSetCollection to generate a contingency table
#using set operations in GSEABase
make_contingency <- function(de_list, GeneSetobj) {


  #make our de list into a simple GeneSet object using GSEABase
  de_geneset <- GeneSet(de_list, setName='1000 DE genes')

  #If we had the full results, we could determine this value without manually setting it
  background_len <- 10000

  #Calculate the values inside the contingency table using set operations
  de_in <- length((de_geneset & GeneSetobj)@geneIds)
  de_notin <- length(setdiff(de_geneset, GeneSetobj)@geneIds)
  set_notde <- length(setdiff(GeneSetobj, de_geneset)@geneIds)
  notin_notde <- background_len - (de_in + de_notin + set_notde)


  #return a matrix of the contingency values
  return(matrix(c(de_in, de_notin, set_notde, notin_notde), nrow=2))


}

contingency_table <- make_contingency(de_genes, hallmarks_gmt[['HALLMARK_ANGIOGENESIS']])
contingency_table
```

We perform the Fisher's Exact test using the built-in R function `fisher.test()`
and view the summarized output by simply calling the variable where we stored
the test results:

```{r fisher test}
fisher_results <- fisher.test(contingency_table, alternative='greater')
fisher_results
```

:::{.box .note}
Specific values of the results can be accessed by the \$notation (i.e.
fisher_results$p.value). The full list of returned values may be found in the R
documentation for `fisher.test()`
:::

Back to our hypothetical example and focusing on the p-value returned of `r
fisher_results$p.value`, we can interpret this as the probability of randomly
obtaining results as or more extreme than what we observed assuming the null
hypothesis that there is no association between differential expression and gene
set membership is true. Based on these results and if this p-value was below our
pre-determined statistical threshold, we could make the conclusion that there is
an *enrichment* or *over-representation* of our differentially expressed genes
from this experiment in the Hallmark Angiogenesis gene set. Relating this back
to the experiment, we might hypothesize that the differences between our two
cell lines might be driving gene expression changes that result in alterations
in genes involved in angiogenesis. This *might* motivate potential further *in
vitro* experiments on these cell lines, including migration and proliferation
assays, that could reveal if this enrichment of angiogenesis genes is reflected
at a phenotypic or functional level.

ORA is a quick and useful way to generate further hypotheses to investigate
specific mechanisms of action or regulation. For example, after identifying a
gene set as being *enriched* or *over-represented*, one could further test the
specific genes in the set by examining their directionality of change or asking
if dependent pathways/networks are also perturbed.

One major limitation of ORA is that it relies on the choice of arbitrary
statistical thresholds to define "interesting" or "differentially expressed"
genes. To reiterate again, p-value thresholds hold no inherent biological
meaning and are subjectively determined. Changing the p-value threshold may
result in dramatic differences in the outcome of ORA. Additionally, expression
datasets may measure tens of thousands of genes in a single experiment, and
filtering by a p-value threshold may discard potentially useful information.

ORA (though often modified with slightly different statistical methodologies) is
implemented in a number of different R packages such as
[topGO](https://bioconductor.org/packages/release/bioc/html/topGO.html) or
various web services including [DAVID](https://david.ncifcrf.gov) and
[enrichR](https://maayanlab.cloud/Enrichr/)

### Rank-based Analysis .

We will refer to the specific method developed by the Broad Institute and UC San
Diego as Gene Set Enrichment Analysis (note the capitalization). This specific
methodology should not be confused with gene set enrichment analysis, which we
use as an umbrella term covering the many statistical methodologies used to 
analyze gene sets. 

[Gene Set Enrichment Analysis] (https://www.gsea-msigdb.org/gsea/index.jsp) was
a method developed to facilitate the biological analysis of genome-wide
expression experiments. It is a rank-based method which utilizes all of the
information from an expression dataset instead of relying on pre-determined
statistical thresholds. In the simplest case, GSEA organizes expression data
into two classes, and ranks all discovered genes by a chosen metric correlating
their expression to these classes. Then, for a pre-defined set of genes, GSEA
tests whether the members of this particular gene set occur more frequently at
the top or bottom of the ranked list or are randomly distributed throughout.
This ranking can be done by a number of different measures, but common ones 
include signal-to-noise ratio, signed log p-value or log fold change estimates. 
These pre-defined gene sets are flexible and may be manually constructed or 
drawn from the many curated databases of gene sets. 

Behind the scenes, GSEA functions by descending through the ranked list and
increasing a cumulative score when a gene is encountered that's contained within
the chosen gene set and decreasing that score when it encounters a gene that is
not within that gene set. This incrementing is weighted to put more emphasis on
genes that are found at the extremes of the ranked list and the final score
taken as the enrichment score (ES) is the maximum of the deviation from zero.
From this score, a p-value is determined by permuting the phenotype labels to
generate a null distribution to compare against. To account for multiple
hypothesis correction, the ES is normalized by the size of the gene set to
create a normalized enrichment score (NES) and the p-values are subjected to
Benjamini-Hochberg correction to generate FDR values for each NES.

To make this explanation more transparent, let us assume that we have performed
a RNAseq experiment to determine which genes are changed when we knock out our
gene of interest in a model cell line.  We have obtained our results, which
consists of a list of all genes discovered in the experiment and associated
statistical measures such as p-value and log fold change estimates. To perform a
basic GSEA, we would sort this list of genes, without respect to significance,
by descending fold change to generate a ranked list where genes that are
upregulated (positive fold change) are at the "top" of our list and genes that
are downregulated (negative fold change) are at the "bottom" of our list. We
would use this ranked list and any number or combination of gene sets to perform
GSEA using one of the available means.

GSEA is available as a java application with a graphical user interface as well
as a command-line utility for use in HPC cluster environments. There also exist
several packages in R that also implement the core GSEA algorithm along with
various changes and refinements to the underlying statistical methodology. In
the following section, we will refer to the usage of one such package, fgsea.
Under most circumstances, GSEA can be run with default parameters but please see
the official documentation for a list of customizable parameters and situations
in which to adjust them.

GSEA will return a list of results for gene sets tested consisting of their
associated enrichment scores, and various statistical measures for a given
ranked input With GSEA results, it is typical to set a permissive FDR threshold
to consider which gene sets are significant. Further inspection of individual
gene sets may be performed by analyzing the leading edge subset for each gene
set. This is the group of genes that occur before and contribute to the maximum
deviation from zero score for a given gene set during the analysis. These genes
are typically interpreted as being the more relevant subset of genes that likely
contribute the most to the enrichment signal, and are typically used in further
investigations comparing leading edge subsets from other gene sets or applying
domain knowledge to predict their regulatory effects and functions with respect
to their annotated biological pathway or gene set. 

GSEA is a highly flexible method that incorporates information from all of the
data generated in HTS experiments. The choice of ranking methods and gene sets
can all be tailored to the specific experiment or question at hand, and it is a
simple yet fast method that can detect more subtle changes in gene networks than
ORA.

:::{.box .warning} 
The interpretation of a GSEA result needs to be made with care. Let's assume
that we have performed GSEA using fold Change as our ranking metric and we find
that the "KEGG Glycolysis" pathway is found to be enriched with a positive
normalized enrichment score and is statistically significant. Alone, this does
not necessarily imply that glycolysis is more active between the conditions of
interest. It simply allows for the conclusion that there is an enrichment of
genes belonging to this gene set among the genes with a positive log fold change
in the experiment. This *might* imply that your experimental conditions are
affecting some underlying transcriptional regulation that results in these genes
being up-regulated. However, it could also be explained by feedback mechanisms,
direct or indirect, from related biological pathways or gene networks. It *does
not* allow you to directly conclude that the glycolysis pathway itself is more
active at a functional level. To make that assertion, you would need to perform
further *in vitro* or *in vivo* functional assays that directly measure the
activity and outputs of glycolysis given your experimental conditions.

It is important to remember that unless specified, gene sets will often contain
a mix of genes with different regulatory activities including both activators
and inhibitors of that pathway. In addition, for many biological pathways, there
are other factors that influence its regulation that may not be apparent at
a gene expression level. In the specific example of metabolic pathways, it is
common for the enzymatic products of subsequent reactions to feedback and
regulate the activity and flux through the pathway or connected pathways. This
regulation is not easily discerned at the gene expression level alone. All of
this is to say that the results from GSEA must be interpreted with care and
integrated with other analyses and knowledge to gain a holistic understanding
of its meaning.
:::
