---
output: html_document
---

```{r data viz setup, include=FALSE}
library(tidyverse)
```

# Data Visualization

Data visualization is a core component of both exploring data and communicating
results to others. The goal of data visualization is to present data in a
graphical way that shows the reader patterns that would not otherwise be
visible. Despite its ubiquity and importance, effective data visualization is
challenging, and while many tools and approaches exist there is no "gold
standard" to follow in any meaningful sense. Rather, an effective visualization
has the following properties:

1. Depicts accurate data
2. Depicts data accurately
3. Shows enough, but not too much, of the data for the viewer to gain insight
4. Is self contained - no additional information (except a caption) is required
   to understand the contents of the figure

Beyond these, a *great* visualization has some additional properties:

5. Exposes patterns in the data not easily observable by other methods
6. Invites the viewer to ask more questions about the data

## Grammar of Graphics {#dv-gg}

The grammar of graphics is a system of rules that describes how data and
graphical aesthetics (e.g. color, size, shape, etc) are combined to form
graphics and plots. First popularized in the book [The Grammar of
Graphics](https://www.google.com/books/edition/_/ZiwLCAAAQBAJ?hl=en) by Leland
Wilkinson and co-authors in 1999, this grammar is a major contribution to the
structural theory of statistical graphics. In 2005, Hadley Wickam wrote an
implementation of the grammar of graphics in R called
[ggplot2](https://ggplot2.tidyverse.org/) (gg stands for grammar of graphics).

Under the grammar of graphics, every plot is the combination of three types of
information: **data**, **geometry**, and **aesthetics**. Data is the data we
wish to plot. Geometry is the type of geometry we wish to use to depict the data
(e.g. circles, squares, lines, etc). Aesthetics connect the data to the geometry
and defines how the data controls the way the selected geometry looks.

A simple example will help to explain. Consider the following made up sample
metadata tibble for a study of subjects who died with Alzheimer's Disease (AD)
and neuropathologically normal controls:

``` {r include=FALSE}
set.seed(1337)
imgsize <- 400**2
ad_metadata <- tibble(
  ID=c(vapply(1:10,function(x) { str_c('A',x)},'a'), vapply(1:10,function(x) { str_c('C',x)},'a')),
  condition=c(rep('AD',10),rep('Control',10)),
tau=c(round(runif(10,0.2,0.9)*imgsize),round(runif(10,0.0,0.4)*imgsize)),
  abeta=round(tau+tau*rnorm(20,0,0.8)),
  iba1=round(tau+tau*rnorm(20,0,0.5)),
  gfap=c(round(runif(10,0.1,0.9)*imgsize),round(runif(10,0.1,0.9)*imgsize)),
  braak_stage=round((tau - min(tau))/(max(tau)-min(tau))*6),
  age_at_death=round(rnorm(20,mean=75,sd=5))+round((tau - min(tau))/(max(tau)-min(tau))*10)-2
) %>% dplyr::select(
    ID,
    age_at_death,
    condition,
    tau,
    abeta,
    iba1,
    gfap,
    braak_stage
)
ad_metadata[ad_metadata<0] <- 0
readr::write_tsv(ad_metadata,"ad_metadata.tsv")
```

``` {r}
ad_metadata
```

For context, [tau protein](https://en.wikipedia.org/wiki/Tau_protein) and
[amyloid beta peptides from the amyloid precursor
protein](https://en.wikipedia.org/wiki/Amyloid_beta)
aggregate into neurofibrillary tangles and A-beta plaques, respectively, the
brains of people with AD. Generally, the amount of both of these pathologies is
associated with more severe disease. [Braak
stage](https://en.wikipedia.org/wiki/Braak_staging) is a neuropathological
assessment of the amount of pathology in a brain that is associated with the
severity of disease, where 0 indicates absence of pathology and 6 with
widespread involvement in multiple brain regions. Aggregation of tau is also a
[consequence of normal aging](https://pubmed.ncbi.nlm.nih.gov/8866683/), so must
accompany neurological symptoms such as dementia to indicate an AD diagnosis
post mortem. Note we have control samples as well as AD.

![Tauopathy: tau protein accumulates in the cell bodies of affected neurons -
Wikipedia](https://upload.wikimedia.org/wikipedia/commons/f/f7/Tauopathy_in_Alzheimer%27s_disease.jpg)

The histology measures `tau`, `abeta`, `iba1`, and `gfap` have been quantified
using digital microscopy, where brain sections are stained with
immunohistochemistry to identify the location and degree of pathology; the
measures in the table are the number of pixels of a 400 x 400 pixel image of a
piece of brain tissue that fluoresce when stained with the corresponding
antibody. Tau and A-beta antibodies are specialized to the types of aggregated
proteins mentioned above and provide a quantification of the level of overall
AD pathology. Ionized calcium binding adaptor molecule 1 (IBA1) is a marker of
activated [microglia](https://en.wikipedia.org/wiki/Microglia), the resident
macrophages of the brain, which is an
indication of neuroinflammation. Glial fibrillary acidic protein (GFAP) is a
marker for activated [astrocytes](https://en.wikipedia.org/wiki/Astrocyte),
specialized cells that derive from the neuron lineage, are critical for
maintaining the [blood brain
barrier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4292164/), and are also
involved in the neuroinflammatory response.

Let's say we wished to visualize the relationship between age at death and the
amount of tau pathology. A scatter plot where each marker is a subject with $x$
and $y$ position corresponding to `age_at_death` and `tau` respectively. The
following R code creates such a plot with ggplot2:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x = age_at_death, y=tau)) +
  geom_point()
```

All ggplot2 plots begin with the `ggplot()` function call, which is passed a
tibble with the **data** to be plotted. We then define  the **aesthetics** are
defined by mapping the x coordinate to the `age_at_death` column and the y
coordinate to the `tau` column with `aes(x = age_at_death, y=tau)`. Finally, the
**geometry** as 'point' with `geom_point()`, meaning marks will be made at pairs
of x,y coordinates. The plot shows what we expect given our knowledge of the
relationship between age and amount of tau; the two look to be positively
correlated.

However, we are not capturing the whole story: we know that there are both AD
and Control subjects in this dataset. How does `condition` relate to this
relationship we see? We can layer on an additional aesthetic of color to add
this information to the plot:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x = age_at_death, y=tau, color=condition)) +
  geom_point()
```

This looks a little clearer, showing that Control subjects generally have both
an earlier age at death and a lower amount of tau pathology. This might be a
problem, however, since if the age distributions of AD and Control groups are
different that might pose a problem with
[confounding](https://en.wikipedia.org/wiki/Confounding). We should investigate
this.

Instead of plotting age at death and tau against each other, we will examine the
distributions of each of these variables for AD and Control samples separately.
We will use the [violin
geometry](https://ggplot2.tidyverse.org/reference/geom_violin.html) with
`geom_violin()` to look at the distributions of `age_at_death`:

``` {r}
ggplot(data=ad_metadata, mapping = aes(x=condition, y=age_at_death)) +
  geom_violin()
```

We can see immediately that there are big differences between the age
distributions of the two groups. This is not ideal, but perhaps we can adjust
for these effects in downstream analyses. We'd like to look at the tau
distributions as well, but it would be nice to have these two plots side by side
in the same plot. To do that, we will use another library called
[patchwork](https://patchwork.data-imaginist.com/), which allows independent
ggplot2 plots to be arranged together with a simple expressive syntax:

``` {r}
library(patchwork)
age_boxplot <- ggplot(data=ad_metadata, mapping = aes(x=condition, y=age_at_death)) +
  geom_boxplot()
tau_boxplot <- ggplot(data=ad_metadata, mapping=aes(x=condition, y=tau)) +
  geom_boxplot()

age_boxplot | tau_boxplot # this puts the plots side by side
```

This confirms our suspicion, and also reveals a serious problem with our
samples: we have strong confounding of tau and age at death between AD and
Control samples. This means that if we look for differences between AD and
Control, we won't know if the difference is due to the amount of tau pathology
or due to age of the subjects. With this sample set, we simply cannot
confidently answer that question. Just a few simple plots alerted us to this
problem; hopefully more expensive datasets have not already been generated for
these samples, so that hopefully different subjects are available that could
avoid this confounding.

This has been a biological data analysis oriented tutorial on plotting meant to
illustrate the principles of the grammar of graphics. Namely, every plot has
**data**, **geometry**, and **aesthetics** that can be independently controlled
to produce many types of plots. Many of these plots have names, like scatter
plots and boxplots, but as you compose different types of geometries and
aesthetics together you may find yourself generating plots that aren't so easily
named.

The next sections of this chapter are a kind of "cook book" of different kinds
plots you can generate with data of different shapes. It is not intended to be
comprehensive, but a helpful guide when you are trying to decide how to
visualize your own datasets.

::: {.box .note}
If you want to go directly to the comprehensive documentation of the many types
of ggplot2 plots, peruse the [R Graph Gallery](https://www.r-graph-gallery.com/)
site.
:::

::: {.box .readmore}
* [ggplot2 - Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html), by Hadley Wickam
* [R for Data Science - Data Vizualization](https://r4ds.had.co.nz/data-visualisation.html)
* [R Graph Gallery](https://www.r-graph-gallery.com/)
* [patchwork package](https://patchwork.data-imaginist.com/)
:::

## Plotting One Dimension

The simplest plots involve plotting a single vector of numbers, or several such
vectors (e.g. for different samples). Each value in the vector typically
corresponds to a category or fixed value, for example the `tau` column from the
example above has pairs of *(ID, tau value)*. The order of these numbers can be
changed, but the vector remains one dimensional or 1-D.

### Bar chart {#dv-bar}

Bar charts map length (i.e. the height or width of a box) to the scalar value of
a number. The difference in visual length can help the viewer notice consistent
patterns in groups of bars, depending on how they are arranged:

``` {r}
ggplot(ad_metadata, mapping = aes(x=ID,y=tau)) +
  geom_bar(stat="identity")
```

Note the `stat="identity"` argument is required because by default `geom_bar`
counts the number of values for each value of x, which in our case is only ever
one. This plot is not particularly helpful, so let's change the fill color of
the bars based on condition:

``` {r}
ggplot(ad_metadata, mapping = aes(x=ID,y=tau,fill=condition)) +
  geom_bar(stat="identity")
```

Slightly better, but maybe we can see even more clearly if we sort our tibble
by tau first. Sorting elements in these 1-D charts is somewhat complicated, and
is explained in the [Reordering 1-D Data Elements] section below.

Bar charts can also plot negative numbers. In the following example, we center
the tau measurements by subtracting the mean from each value before plotting:

``` {r}
mutate(ad_metadata, tau_centered=(tau - mean(tau))) %>%
  ggplot(mapping = aes(x=ID, y=tau_centered, fill=condition)) +
  geom_bar(stat="identity")
```

### Lollipop plots

Similar to [bar charts](#dv-bar), so-called "lollipop plots" replace the bar
with a line segment and a circle. The length of the line segment is proportional
to the magnitude of the number, and the point marks the length of the segment as
a height on the y or length on the x axis, depending on orientation.

``` {r}
ggplot(ad_metadata) +
  geom_point(mapping=aes(x=ID, y=tau)) +
  geom_segment(mapping=aes(x=ID, xend=ID, y=0, yend=tau))
```

Note that `aes()` mappings can be made on the `ggplot()` object or on each
individual geometry function call, to specify different mappings based on
geometry.

Area/Stacked Area charts

### Parallel Coordinate plots

## Visualizing Distributions

The distribution is one of the most important properties of a set of numbers.
A distribution describes the general "shape" of the numbers, i.e. what is the
relative frequency of the values, or ranges of values, within the data.
Understanding the distribution of a data set is critical when choosing methods
to apply, since many methods are only appropriate when data is distributed in
certain ways, e.g. [linear
regression](https://en.wikipedia.org/wiki/Linear_regression) assumes the
response variable is [normally
distributed](https://en.wikipedia.org/wiki/Normal_distribution), otherwise the
result of the model cannot be interpreted properly. Often, we don't know how our
data are distributed when we obtain it and so we must examine the distribution
empirically. The visualizations in this section are all used for the purpose of
depicting the distribution of a set of numbers.

### Histogram

The most common way to plot the distribution of a 1-D set of data is the
[histogram](https://en.wikipedia.org/wiki/Histogram). The histogram divides up
the range of a dataset from minimum to maximum into bins usually of the same
width and tabulates the number of values that fall within each bin. Below is a
histogram of our `age_at_death` measurement for all samples:

``` {r}
ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death))
```

Note that the histogram does not look very complete, because there are only 20
values in our data. We can mitigate this somewhat by increasing the number of
bins the data range is divided into:

``` {r}
ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death),bins=10)
```

This is a little bit better, but there are still some bins (76-79, 84-87) that
have no values. Compare this to the following synthetic dataset of 1000
normally distributed values:

``` {r}
tibble(
  x=rnorm(1000)
) %>%
  ggplot() +
  geom_histogram(aes(x=x))
```

For distributions with a small number of samples, histograms might not be the
best visualization. We will continue with synthetic normally distributed dataset
for the remaining examples.

ggplot allows you to easily plot multiple distributions on the same plot:

``` {r}
tibble(
  x=c(rnorm(1000),rnorm(1000,mean=4)),
  type=c(rep('A',1000),rep('B',1000))
) %>%
  ggplot(aes(x=x,fill=type)) +
  geom_histogram(bins=30, alpha=0.6, position="identity")
```

The `alpha=0.6, position="identity"` arguments makes the bars partially
transparent so you can see the overlap more clearly.

::: {.box .readmore}
* [ggplot2 geom_histogram reference](https://ggplot2.tidyverse.org/reference/geom_histogram.html?q=histogram)
* [R Graph Gallery - histograms](https://www.r-graph-gallery.com/histogram.html)
:::

### Density

Another way to describe a distribution is with a [density
plot](https://www.r-graph-gallery.com/density-plot.html). Instead of binning the
values into intervals and drawing bars with height proportional to the number of
values in each bin, a density plot draws a smoothly interpolated line that
approximates the distribution instead. A key difference between a histogram and
a density plot is the density plot is always normalized so the integral under
the curve is approximately 1, whereas a histogram may be either counts or, if
the counts in each bin are divided by the total number of data points, a
proportion.

Compare the histogram and density plots
of the  `age_at_death` variable from our example tibble:

``` {r}
library(patchwork)
hist_g <- ggplot(ad_metadata) +
  geom_histogram(mapping=aes(x=age_at_death),bins=30)
density_g <- ggplot(ad_metadata) +
  geom_density(mapping=aes(x=age_at_death),fill="#c9a13daa")

hist_g | density_g
```

Notice the overall shape of the two distributions is similar, with the highest
values in both around age 77. The density plot is a smoother representation of
a histogram, but its accuracy is still highly sensitive to the number of
measurements used to construct it. Compare the histogram and density plots of
two sets of 1000 normally distributed samples with different means:

``` {r}
library(patchwork)
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,mean=4)),
  type=c(rep('A',1000),rep('B',1000))
)
hist_g <- ggplot(normal_samples) +
  geom_histogram(
    mapping=aes(x=x,fill=type),
    alpha=0.6,
    position="identity",
    bins=30
)
density_g <- ggplot(normal_samples) +
  geom_density(
    mapping=aes(x=x,fill=type),
    alpha=0.6,
    position="identity"
  )

hist_g | density_g
```

Again the two types of plots depict similar distributions, although they are
different enough to possibly suggest different interpretations. In general,
density plots might be preferable over histograms if the data are noisy or
sparse in that they produce cleaner plots, but potentially at the expense of
accuracy when the number of samples is low.

::: {.box .readmore}
* [ggplot2 geom_density reference](https://ggplot2.tidyverse.org/reference/geom_density.html?q=density)
* [R Graph Gallery - density plots](https://www.r-graph-gallery.com/density-plot.html)
:::

### Boxplot

Box plots, or box and whisker plots, are extremely common when used to describe
distributions. Below is a boxplot of age at death divided by condition:

``` {r}
ggplot(ad_metadata) +
  geom_boxplot(mapping=aes(x=condition,y=age_at_death))
```

Boxplots are drawn assuming the data are
[unimodal](https://en.wikipedia.org/wiki/Unimodality) (i.e. shaped like a hill,
possibly slanted to one side or the other), where the extents of the box
represent the 1st and 3rd quartile of the data, the central line is the median,
the whiskers are drawn as 1.5 times the value outside the 1st and 3rd quartiles.
Sometimes individual values more extreme than the whiskers are drawn
individually to identify them as outliers.

![Boxplot anatomy. IQR stands for "inner quartile range", the distance between
the 1st and 3rd quartile - Wikipedia](https://upload.wikimedia.org/wikipedia/commons/1/1a/Boxplot_vs_PDF.svg)

However, boxplots have some significant shortcomings. Primarily, the rectangle
of the inner quartile range does not describe the actual distribution of the
samples within it. Although the median can give a sense of skewness, if the
data are not unimodal this may be misleading. Consider the following
distributions plotted as boxplots or as violin plots (described in the next
section):

``` {r}
library(patchwork)
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,4),rnorm(1000,2,3)),
  type=c(rep('A',2000),rep('B',1000))
)
g <- ggplot(normal_samples, aes(x=type,y=x,fill=type))
boxplot_g <- g +  geom_boxplot()
violin_g <- g + geom_violin()

boxplot_g | violin_g
```

The two distributions look almost identical in the boxplot figure; however they
are dramatically different when visualized using a method like a violin plot
where the contours of the entire distribution are depicted. Unless you are
certain that your data are unimodal, one of the other distribution visualization
methods in this section will likely more accurately depict your data than a
boxplot.

::: {.box .readmore}
* [ggplot2 geom_boxplot reference](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)
* [R Graph Gallery - boxplots](https://www.r-graph-gallery.com/boxplot.html)
:::

### Violin plot

As seen in the last section, a [violin
plot](https://ggplot2.tidyverse.org/reference/geom_violin.html) is another way
to depict a distribution by producing a shape where the width is proportional to
the value along the x or y axis, depending on orientation. The "violin" shape is
similar in principle to a histogram or a density plot, in that it describes the
contour of all the data in the distribution, not just the quantiles and extents,
as in a box plot. Below is a violin plot of the `tau` measures from our example
tibble:

``` {r}
ggplot(ad_metadata) +
  geom_violin(aes(x=condition,y=tau,fill=condition))
```

The violin plot is both more and less descriptive than a boxplot; it does depict
the entire distribution of the data, but also doesn't include features like
median by default.

::: {.box .readmore}
* [ggplot2 geom_violin reference](https://ggplot2.tidyverse.org/reference/geom_violin.html)
* [R Graph Gallery - violin plots](https://www.r-graph-gallery.com/violin.html)
:::

### Beeswarm plot

The [beeswarm plot](https://r-charts.com/distribution/ggbeeswarm/) is similar to
a violin plot, but instead of plotting the contours of the data, it plots the
data itself as points like in a scatter plot. The individual values of the
distribution are organized vertically and spaced such that the points don't
overlap. In this plot, the distribution of age at death is plotted for each kind
of sample and the markers are colored by the amount of tau:

``` {r}
library(ggbeeswarm)
ggplot(ad_metadata) +
  geom_beeswarm(aes(x=condition,y=age_at_death,color=condition),cex=2,size=2)
```

We may not have noticed before that our AD samples have a big gap in ages
between 74 and 81; since the beeswarm plot displays all the data, we can see it
easily here.

Beeswarm plots are typically only useful when the number of values is within a
range; not too many and not too few. The example above is close to having too
few values per group for this plot to be useful, but consider the following with
too many samples:

``` {r}
normal_samples <- tibble(
  x=c(rnorm(1000),rnorm(1000,4),rnorm(1000,2,3)),
  type=c(rep('A',2000),rep('B',1000))
)
ggplot(normal_samples, aes(x=type,y=x,color=type)) +
  geom_beeswarm()
```

This plot likely has too many samples to be the right choice (it's also ugly),
but it does give an idea of the distribution of the data.

In the previous examples the markers for each group also determined the color of
the group. This makes the chart a bit easier to read and more pleasing to the
eye, but is technically redundant. You can use however profitably however to
color markers by some other value that might be of interest. Consider this final
example where markers are colored by another randomly generated variable:

``` {r}
normal_samples <- tibble(
  x=c(rnorm(100),rnorm(100,4),rnorm(100,2,3)),
  type=c(rep('A',200),rep('B',100)),
  category=sample(c('healthy','disease'),300,replace=TRUE)
)
ggplot(normal_samples, aes(x=type,y=x,color=category)) +
  geom_beeswarm()
```

We are now effectively visualizing three dimensions which may provide insight
into the data.

::: {.box .readmore}
* [beeswarm package reference](https://r-charts.com/distribution/ggbeeswarm/)
:::

### Ridgeline

If you have many non-trivial distributions that you would like the user to
compare, a good option is a [ridgeline
chart](https://www.r-graph-gallery.com/294-basic-ridgeline-plot.html). The
ridgeline plot is simply multiple density plots drawn for different variables
within the same plot. Like the beeswarm plot, ridgeline plots are provided by
another package outside ggplot2.

``` {r}
library(ggridges)

tibble(
  x=c(rnorm(100),rnorm(100,4),rnorm(100,2,3)),
  type=c(rep('A',200),rep('B',100)),
) %>%
  ggplot(aes(y=type,x=x,fill=type)) +
  geom_density_ridges()
```

Many distributions may be plotted:

``` {r}
tibble(
  x=rnorm(10000,mean=runif(10,1,10),sd=runif(2,1,4)),
  type=rep(c("A","B","C","D","E","F","G","H","I","J"),1000)
) %>%
  ggplot(aes(y=type,x=x,fill=type)) +
  geom_density_ridges(alpha=0.6,position="identity")
```

::: {.box .readmore}
* [R Graph Gallery - ridgeline plots](https://www.r-graph-gallery.com/294-basic-ridgeline-plot.html)
* [ggridges package vignette](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html)
* [ggridges package index on CRAN](https://cran.r-project.org/web/packages/ggridges/index.html)
:::

## Plotting Two or More Dimensions

### Scatter plot

### Line plot

``` {r}
ggplot(ad_metadata,mapping=aes(x=ID, y=tau)) +
  geom_line()
```

### Bubble plot

### Heatmaps

## Other Kind of Plots

Chord

Sankey

Dendrogram

## Network Viz {#ds-network-viz}

## Tips and Tricks

Reordering 1-D Data Elements

``` {r}
ad_metadata %>%
arrange(tau) %>%
ggplot(mapping = aes(x=ID,y=tau,fill=condition)) +
  geom_bar(stat="identity")
```

But wait, our bars are in the same order as in the previous plot. Sorting our
data prior to plotting did not have the desired effect. The reason is that
although the `ID` column is a character string in our original tibble, ggplot
converts it into a [factor](#prog-factor) prior to plotting. Recall that a
factor is a numeric type composed of sequential integers, each of which having a
character label associated with it.


Confidence Intervals

Annotations

## Multiple Plots

### Facet wrapping

## Responsible Plotting {#dv-resp}

"Good plots empower us to ask good questions." - Alberto Cairo, How Charts Lie

## Publication Ready Plots

::: {.box .readmore}
* [ggpubr - publication ready plotting with ggplot](https://rpkgs.datanovia.com/ggpubr/)
:::
