---
title: "BF591 - R for Biological Sciences"
bibliography: references.bib
link-citations: true
---

```{r include=FALSE}
# github pages doesn't host files/directories that start with "_", change it for the whole book
knitr::opts_chunk$set(fig.path = 'main_files/')
# always turn off warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# cache by default
knitr::opts_chunk$set(cache = TRUE)
```

# R for Biological Sciences {.unnumbered}

**Semester: Fall 2024**

**Location: EPC 209**

**Time: M/W 10:10 - 11:55**

```{r, echo=F, results="asis"}
cat(paste("Site last updated",format(Sys.time(), "%a %b %d %Y")))
```

Contents:

-   [Course Schedule]
-   [Instructors]
-   [Office Hours]
-   [Course Values and Policies]

This course introduces the R programming language through the lens of
practitioners in the biological sciences, particularly biology and
bioinformatics. Key concepts and patterns of the language are covered,
including:

-   RStudio
-   Data wrangling with tidyverse
-   Data visualization with ggplot
-   Essential biological data shapes and formats
-   Core bioconductor packages
-   Basic data exploration, including elementary statistical modeling
    and summarization
-   Elementary Data Science concepts
-   "Toolifying" R scripts
-   Communicating R code and results with RMarkdown
-   Building R packages and unit testing strategies
-   Building interactive tools with RShiny

About 1/3 of the materials are inspired by the online textbook [R for
Data Science](https://r4ds.had.co.nz), while the rest has been developed
by practicing bioinformaticians based on their experiences.

Weekly programming assignments will help students apply these techniques
to realistic problems involving analysis and visualization of biological
data. Students will be introduced to a unit testing paradigm that will
help them write correct code and deposit all their code into github for
evaluation. Students will implement an end-to-end project that begins
with one of a set of provided datasets, implements a set of data
summarization and exploration operations on it, and allows interaction
with an RShiny app.

Relevant links (zoom and textbook) are available on Blackboard as a reference.
The blackboard site will only be used for announcements and to host links. 

## Course Schedule {.unnumbered}

**Key:**

-   Follow [Week N](#class-outlines) links for detailed list of topic
    sections.
-   Assignments are assigned and due on the first day of class each week
    unless mentioned otherwise

| Week                                              | Dates         | Topics                                                                                                                                                  | Assignment                       |
| ------------------------------------------------- | ------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------- | -------------------------------- |
| [Week 1] \| [slides](slides/week_1_slides.html)   | 9/4 & 9/9     | - Preliminaries<br>- Data in Biology<br>- R Programming Basics                                                                                          | [Assignment 0]                   |
| [Week 2] \| [slides](slides/week_2_slides.html)   | 9/11 & 9/16   | - EngineeRing: Unit Testing<br>- Assignment Structure<br>- Data Wrangling & Tidyverse Basics                                                            | [Base R]                         |
| [Week 3] \| [slides](slides/week_3_slides.html)   | 9/18 & 9/23   | - R in Biology<br>- Bio: Bioconductor Basics                                                                                                            | [Tidyverse Basics]               |
| [Week 4] \| [slides](slides/week_4_slides.html)   | 9/25 & 9/30   | - Data Viz: Grammar of Graphics<br>- Bio: Gene Expression<br>- Bio: Microarrays<br>- Data Sci: Data Modeling<br>- Bio: Differential Expression          | [Bioinformatics Basics]          |
| [Week 5] \| [slides](slides/week_5_slides.html)   | 10/2 & 10/7   | - Data Sci: PCA & Clustering<br>- Data Viz: Heatmaps & Dendrograms<br>- R Programming: Structures and Iteration                                         | [Data Science Basics]            |
| [Week 6] \| [slides](slides/week_6_slides.html)   | 10/9 & 10/15  | - Bio: High Throughput Sequencing<br>- Bio: RNAseq<br>- Bio: Count Data<br>- Bio: RNAseq Differential Expression<br>- Bio: Gene Set Enrichment Analysis | [Counts Analysis]                |
| [Week 7] \| [slides](slides/week_7_slides.html)   | 10/16 & 10/21 | - Data Sci: Distributions<br>- Data Sci: Statistical Tests<br>- R Programming: Styles and Conventions                                                   | [Differential Expression Part 1] |
| [Week 8] \| [slides](slides/week_8_slides.html)   | 10/23 & 10/28 | - Data Vis: Plot Week<br>- Data Vis: Responsible Plotting                                                                                               | [Differential Expression Part 2] |
| [Week 9] \| [slides](slides/week_9_slides.html)   | 10/30 & 11/4  | - Rshiny                                                                                                                                                | [RShiny Basics]                  |
| [Week 10] \| [slides](slides/week_10_slides.html) | 11/6 & 11/11  | - engineeRing                                                                                                                                           | [RShiny Basics]                  |
| [Week 11]                                         | 11/13 & 11/18 | - Project Work                                                                                                                                          | [Final Project]                  |
| [Week 12]                                         | 11/20 & 11/25 | - Professional Development Workshops                                                                                                                    |                                  |
| [Week 13]                                         | 12/2 & 12/4   | - Professional Development Workshops                                                                                                                    |                                  |
| [Week 14]                                         | 12/9          | - Course Feedback                                                                                                                                       |                                  |

## Instructors {.unnumbered}

Primary instructors: 

Joey Orofino (jorofino AT bu DOT edu)

Adam Labadorf (labadorf AT bu DOT edu)

TAs:

Xudong Han

Eric Palanques Tost

## Office Hours {.unnumbered}

Regular: Tuesdays 11-12pm, LSE101, Zoom link on Blackboard

By appointment: Email me (Joey)

TAs: TBA

## Course Values and Policies {.unnumbered}

**Everyone is welcome.** Every background, race, color, creed, religion,
ethnic origin, age, sex, sexual orientation, gender identity,
nationality is welcome and celebrated in this course. Everyone deserves
respect, patience, and kindness. Disrespectful language, discrimination,
or harassment of any kind are not tolerated, and may result in removal
from class or the University. This is not merely [BU
policy](http://www.bu.edu/policies/policy-category/harassment-discrimination/).
The instructors deem these principles to be inviolable human rights.
Students should feel safe reporting any and all instances of
discrimination or harassment to the instructor, to any of the
Bioinformatics Program leadership, or the [BU Equal Opportunity
Office](http://www.bu.edu/eoo/).

**Everyone brings value.** Each of us brings unique experiences, skills,
and creativity to this course. Our diversity is our greatest asset.

**Collaboration is highly encouraged.** All students are encouraged to
work together and seek out any and all available resources when
completing projects in all aspects of the course, including sharing
coding ideas and strategies with each other as well as those found on
the internet. Any and all available resources may be brought to bear.
However, consistent with BU policy, the bulk of your code and your final
reports should be written in your own words and represent your own work
and understanding of the material. Copying/pasting large sections of
code is not acceptable and will be investigated as cheating (we check).

**A safe space for dissent.** For complex topics such as those covered
in this class, there is seldom one correct answer, approach, or
solution. Disagreement fosters innovation. All in the course, including
students and TAs, are encouraged to express constructive criticism and
alternative ideas on any aspect of the content.

**We are always learning.** Our knowledge and understanding is always
incomplete. Even experts are fallible. The bioinformatics field evolves
rapidly, and Rome was not built in a day. Be kind to yourself and to
others. You are always smarter and more knowledgeable today than you
were yesterday.

## Grading {.unnumbered}

Grading will be based on the 7 roughly weekly assignments and the final project.
Each assignment is 5% of your total grade (35% total), and the final project is
60% of your grade. The remaining 5% is for class attendance / participation.

## Absences, missed classes, and extensions {.unnumbered}

You should always prioritize your physical, emotional and mental health. BU
offers a number of resources through Student Health Services and I encourage you
to explore them if you feel you need to talk to someone. I am also here to
listen without judgement if needed and can help with accessing the resources
available.

If you need to miss extended class time or require extra time on an assignment
because of personal matters, please just inform me and I will work with you
whenever you are back to catch you up on the material and find an arrangement
that accommodates your needs. You never need to disclose to me any private
matters if you are not comfortable doing so.


## Acknowledgements & Contributions {.unnumbered}

These materials would not have been possible without the contributions
of Dakota Hawkins, Vanessa Li, Taylor Falk, and Mae Rose Gott, and Joey
Orofino.

Former valiant and (probably still) attractive TAs:

## 2023 {.unnumbered}

-   **Regan Conrad** (BU BF PhD Candidate)
-   **Aubrey (Brie) Odom-Mabey** (BU BF PhD Candidate)

### 2022 {.unnumbered}

-   **Taylor Falk** (BU BF MS Alumnus '21) is a Bioinformatics Developer
    working with the VA PTSD Brain Bank, developing infrastructure to
    support data generated out of our brains. He is championing the
    assignment strategy and will be available to help with issues.
-   **Mae Rose Gott** (BU BF MS Alumna '21) is a Research Staff member
    working on a number of different projects across many different
    areas. She will be helping organize the course materials as we go
    forward.
-   **Vanessa Li** (BU BF PhD Candidate) is a PhD candidate in Dr.
    Stefano Monti's lab in Computational Biomedicine. She will primarily
    be helping with grading of your assignments.
-   **Joey Orofino** (BU BF MS Alumnus '15) is a Research Scientist
    working on identifying small RNA based biomarkers of Parkinson's
    Disease
