---
title: "Autograding Report"
output: html_document
params:
  status: "bf591-assignment-2_status.csv"
  tests: "bf591-assignment-2_tests.csv"
---

```{r Libraries and setup, echo=FALSE, results='hide', message=FALSE}
suppressPackageStartupMessages({
  library(tidyverse, quietly = T)
  library(kableExtra)
})
commandArgs(TRUE)
foldername <- strsplit(params$status, "\\.")[[1]][1]
foldername <- strsplit(foldername, "_")[[1]][1]
```
This script was generated using `autograde.sh` located in:
```{bash, echo=FALSE}
pwd
```
This was used to test **`r paste0(foldername, "/")`** on `r date()`.

## Student Status
Below are the student folders found within the testing directory. If a student's directory contains a folder called **`r paste0(foldername, "/")`** then the assingment folder will read as `TRUE`. If, within that folder, there is a file called `test_main.R` then the testing file is present.
```{r Draw status, echo=FALSE}
statuscsv <- read_csv(params$status, show_col_types = F)
statuscsv$student_folder <- str_replace_all(statuscsv$student_folder, 
                                            pattern = "[\\./]", 
                                            replacement = "")
statuscsv %>%
  mutate(folder_exists = cell_spec(folder_exists, 
                                   background = ifelse(folder_exists, "lightgreen","pink")),
         test_exists = cell_spec(test_exists, 
                                 align = "center",
                                 background = ifelse(test_exists, "lightgreen", "pink"))) -> 
  statuscsv 

colnames(statuscsv) <- c("Student folder", 
                         "Assignment Folder Exists", 
                         "test_main.R Exists")
statuscsv %>%
  kable(escape = FALSE, align = "c") %>%
  kable_styling(full_width = TRUE)
```

## Testing Results
Based on the tests available in `test_main.R`, each student's code will output whether a test has passed, failed, been skipped, or thrown a warning.

```{r Plot test results, echo=FALSE}
tests <- read_csv(params$tests, show_col_types = F)
num_tests <- nchar(tests$test_result[1])
tests <- cbind(tests, str_split_fixed(tests$test_result, "", n = num_tests))
tests$percent <- apply(tests, 
                       function(x) { length(which(x == "."))/ num_tests}, 
                       MARGIN = 1)

add_rect_row <- function(ylevel, status_arr) {
  data.frame(xmin=0:(length(status_arr)-1),
             xmax=1:length(status_arr),
             ymin=rep(ylevel+0.15, 4),
             ymax=rep(ylevel+0.85, 4),
             fill=status_arr) %>%
    return()
}

plotdf <- data.frame(xmin=NULL, xmax=NULL, ymin=NULL, ymax=NULL, fill=NULL)
for (row in 1:dim(tests)[1]) {
  plotdf <- rbind(plotdf, add_rect_row(row-1, 
                                       as.character(tests[row, 4:dim(tests)[2]-1])))
}

ggplot(plotdf, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax)) +
  geom_rect(aes(fill=fill, color="black")) +
  scale_fill_manual("Test Status", 
                    values=c("."="lightgreen", "E"="coral1", "S"="cyan3", "W"="orange"),
                    labels=c("Pass", "Error", "Skipped", "Warning")) +
  scale_colour_manual(values=c("."="lightgreen", "E"="coral1", "S"="cyan3", "W"="orange"), 
                      guide="none") +
  scale_x_continuous("Test",
                     breaks=0.35:(dim(tests)[2]-3),
                     labels=1:(dim(tests)[2]-3)) +
  scale_y_continuous("Student Folder",
                     breaks=0.5:(dim(tests)[1]),
                     labels=tests$student_dir) +
  theme(axis.text.y = element_text(angle=30),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

```

